Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-05-18 papers/paper-eccentricity/paper-eccentricity.tex
More writing, flesh out some other sections

\textit{Software:} \ketju{} \citep{rantala17}, \gadget{} \citep{springel22}, NumPy \citep{numpy}, SciPy \citep{scipy}, Matplotlib \citep{matplotlib}, pygad \citep{rottgers20}.

>>>>

\textit{Software:} \ketju{} \citep{rantala2017}, \gadget{} \citep{springel2022}, NumPy \citep{harris2020}, SciPy \citep{virtanen2020}, Matplotlib \citep{hunter2007}, pygad \citep{rottgers2020}.

### 2023-05-18 papers/paper-eccentricity/sections/discuss.tex
More writing, flesh out some other sections

>>>>

We present in \autoref{sec:results} evidence for a tight correlation between the first hard-scattering angle $\theta_\mathrm{defl}$ and the hard SMBH binary eccentricity. 
With the inclusion of the toy model, which provides a resolution-free model of the SMBH scattering process, we may draw two key interpretations of the results.

First, the numerical simulations performed with \ketju{} are deterministic for a given set of initial conditions, given the same computing architecture and numerical parameters (e.g. the number of computing cores used).
The scatter we observe in $\theta_\mathrm{defl}$ is thus a manifestation of Brownian motion induced by the phase space discretisation and other numerical errors, such as floating point round-off errors, denoted $\sigma_\mathrm{sim}$. 
Thus, the numerical simulations present an instance where there is zero uncertainty in the initial conditions of the system, however there is error in the numerical evolution of the system.

Second, in the real Universe there is no numerical error affecting the system evolution.
There is however uncertainty in the initial conditions of the system, whereby the chance of two galaxy merger events having the exact same initial configuration is practically zero.
We denote this uncertainty as $\sigma_\mathrm{true}$.
Intuitively, we expect that $\sigma_\mathrm{sim} > \sigma_\mathrm{real}$: in other words, $\sigma_\mathrm{sim}$ provides an upper estimate on the uncertainty we can expect from galaxy mergers in the real Universe. 
In particular, processes that perturb the impact parameter associated with the hard scattering of a SMBH pair will manifest as an effect in the resulting binary eccentricity.

### 2023-05-18 papers/paper-eccentricity/sections/numerical.tex
More writing, flesh out some other sections

\begin{itemize}
    \item describe toy model
    \item describe galaxy ICs: 0.5 Dehnen spheres
    \item describe merger orbit: low and high e
\end{itemize}

\subsection{Galaxy ICs}\label{ssec:gics}
We construct our galaxy models to match those of \citet{gualandris22}.
Each galaxy is represented as a stellar-only Dehnen sphere \citep{dehnen93} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed with zero initial velocity.
We test \drafting{X} different mass resolutions by sampling the stellar phase space distribution with varying number of stellar particles: \drafting{list resolutions here}.

\subsection{Merger ICs}
We construct isolated merger initial conditions by placing two galaxies on an elliptical orbit with varying values of eccentricity, at a fixed initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
We follow the same orbital configurations as \citet{gualandris22}, with three additional orbit. 
The chosen eccentricities $e$ are 0.5, 0.7, 0.9, 0.95, 0.97, and 0.99.
We ensure that the radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris22}.
For each orbital configuration, we run ten realisations, to account for Brownian motion effects of the SMBH binary caused by the discretised phase space. 
Each orbital geometry is simulated using our fiducial particle resolution of $M_\bullet/m_\star = 10^4$.
Additionally, we run the $e_0=0.9$ and $e_0=0.99$ orbits at varying mass resolutions, using the galaxy ICs at the mass resolutions described in \autoref{ssec:gics}.

@@

The first method follows \citet{nasim20}, and is an inverse problem. 
For each simulation in a given set, we determine the mean eccentricity $e_\mathrm{h}$ over five orbital periods centred on the orbit within which the SMBH binary has become hard.
The binary hardening radius $a_\mathrm{h}$ is defined as \citep[e.g.][]{merritt13}
\begin{equation}\label{eq:ahard}
    a_\mathrm{h} = \frac{q}{1+q} \frac{r_\mathrm{h}}{4},
\end{equation}
where $r_\mathrm{h}=r(m<2M_{\bullet,1})$ is the influence radius and $q = \{q \in \mathbb{R} | 0<q \leq 1\}$ is the mass ratio between the SMBHs.
Under the assumption that each calculated mean eccentricity is a sample of some common distribution, we may (assuming that the distribution is Gaussian) obtain an estimate of the standard deviation of this distribution by taking the standard deviation of the collection of sample eccentricity means.

>>>>

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{rantala2017}, implemented in \gadget{}.
For details about the implementation of \ketju{}, we point the reader to \drafting{Mannerkoski in prep}.

\subsection{Galaxy ICs}\label{ssec:gics}
We construct our galaxy models to match those of \citet{gualandris2022}.
Each galaxy is represented as a stellar-only Dehnen sphere \citep{dehnen1993} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed with zero initial velocity.
We test five different mass resolutions by sampling the stellar phase space distribution with varying number of stellar particles: $N_\star = \{2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6\}$.

\subsection{Merger ICs}
We construct isolated merger initial conditions by placing two galaxies on an elliptical orbit with varying values of eccentricity, at a fixed initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
We follow the same orbital configurations as \citet{gualandris2022}, with three additional orbit. 
The chosen eccentricities $e$ are 0.5, 0.7, 0.9, 0.95, 0.97, and 0.99.
We ensure that the radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris2022}.
For each orbital configuration, we run ten realisations, to account for Brownian motion effects of the SMBH binary caused by the discretised phase space. 
Each orbital geometry is simulated using our fiducial particle resolution of $M_\bullet/m_\star = 10^4$.
Additionally, we run the $e_0=0.9$ and $e_0=0.99$ orbits at varying mass resolutions, using the galaxy ICs at the mass resolutions described in \autoref{ssec:gics}.

@@

The first method follows \citet{nasim2020}, and is an inverse problem. 
For each simulation in a given set, we determine the mean eccentricity $e_\mathrm{h}$ over five orbital periods centred on the orbit within which the SMBH binary has become hard.
The binary hardening radius $a_\mathrm{h}$ is defined as \citep[e.g.][]{merritt2013}
\begin{equation}\label{eq:ahard}
    a_\mathrm{h} = \frac{q}{1+q} \frac{r_\mathrm{h}}{4},
\end{equation}
where $r_\mathrm{h}=r(m<2M_{\bullet,1})$ is the influence radius and $q = \{q \in \mathbb{R} | 0<q \leq 1\}$ is the mass ratio between the SMBHs.
Under the assumption that each calculated mean eccentricity is a sample of some common distribution, we may (assuming that the distribution is Gaussian) obtain an estimate of the standard deviation of this distribution by taking the standard deviation of the collection of sample eccentricity means.

### 2023-05-18 papers/paper-eccentricity/sections/results.tex
More writing, flesh out some other sections

\section{Results}

\begin{itemize}
    \item Show that brownian motion affects impact parameter of the final passage before BHs bound --> eccentricity set
\end{itemize}

@@

\subsection{Eccentricity scatter in low-e orbits}
\begin{itemize}
    \item Show that impact parameters not sensitive
\end{itemize}
We show the dependence of eccentricity standard deviation on mass resolution for the $e_0=0.90$ orbit in \autoref{fig:convergence} with purple markers. 
Taking the full set of ten simulations for this set (circle markers), immediately apparent is the lack of convergence of $\sigma_e$, with $5\times10^3$ and $1\times10^4$ mass resolutions demonstrating an increased scatter compared to the lowest mass resolution of $2.5\times10^3$.
The peculiar values of $\sigma_e$ arises due to the assumed Gaussian distribution of the eccentricity between simulations.
Inspecting the orbital parameters for all simulations that form the $e_0=0.90$ suite, it is found that for resolutions of $5\times10^3$ and $1\times10^4$ the eccentricity clusters around a value of $e_\mathrm{h}\simeq0.90$, except for one simulation at each resolution which has a value of $e_\mathrm{h}<0.4$.
This one outlier for each of the two resolutions artificially inflates the measured eccentricity variance, and by applying a $2.5\sigma$-cut in eccentricity we recover the $1/\sqrt{M_\bullet/m_\star}\propto 1/\sqrt{N_\star}$ scaling of $\sigma_e$ as in \citet{nasim20} (square markers), apart from the highest mass resolution set of $M_\bullet/m_\star=4\times10^4$.
However, the eccentricity values beyond $2.5\sigma$ from the mean are each a valid measurement from the simulation, and highlight the presence of stochasticity in eccentricity for these idealised simulation set-ups. 
The cause of these low eccentricity values is discussed in \autoref{ssec:toy}.

\subsection{Eccentricity scatter in high-e orbits}
\begin{itemize}
    \item show that impact parameters sensitive
\end{itemize}
We show the dependence of eccentricity standard deviation on mass resolution for the $e_0=0.99$ orbit in \autoref{fig:convergence} with blue markers, where the marker symbols are consistent with those for the $e_0=0.90$ suite. 
We find that the scaling of $\sigma_e$ is steeper than the $1/\sqrt{M_\bullet/m_\star}\propto1/\sqrt{N_\star}$ relation described in \citet{nasim20}.
We attribute this dramatic decrease in $\sigma_e$ to an inflated value of $\sigma_e$ at low mass resolution: there is increased scatter in eccentricity at low mass resolutions for very radial initial orbits of the SMBH binary. 
This observation is highlighted in \autoref{fig:nasim_e_comp}.
As the majority of major galaxy mergers in the real Universe are expected to be radial \citep[e.g.][]{khochfar06}, understanding if the origin of the scatter in eccentricity is a numerical or a physical effect is of paramount importance, and will have direct implications for the predictions of SMBH merger rates by ongoing and future GW detection missions.

@@

After specifying the initial position and velocity, as well as the other parameters of the model,
the equation of motion can be solved using standard numerical integration techniques.
We use the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{scipy}
for this purpose.
To mimic the nearly linearly plunging orbit in the high-eccentricity merger orbit simulations,
we specify the initial conditions confined to the $xz$-plane as 
\begin{gather}
\vb{x}_0 = (r_0, b)\\
\dot{\vb{x}}_0 = (-v_0, 0).
\end{gather}
We fix $r_0 = 100\,\mathrm{pc}$,
and use model parameters
$\rho = 400 \,\Msun\,\mathrm{pc}^{-3}$,
$\sigma=200\,\mathrm{km\,s^{-1}}$,
$M_\bullet=10^8\,\Msun$,
and
$e_\mathrm{s}=0.85$,
which match fairly well with the central properties of the merging galaxy when
the SMBHs are becoming bound in the $e_0=0.99$ merger orbit simulations.
\drafting{(Check the params)}

\drafting{This right place for this section?}
To investigate the effect of the deflection angle, and by extension the impact parameter, on the final SMBH binary eccentricity, we determine the deflection angle at each pericentre passage before the SMBHs become bound \citep{binney08}:
\begin{equation}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, and $L$ is the magnitude of the SMBH binary angular momentum vector and $E$ the binary orbital energy at the time of the pericentre passage.
We set as the first `hard-scattering' angle the first angle that exceeds an orbit-dependent threshold: $\theta_{\mathrm{defl,min}}=90$ for $e_0=0.99$ and $\theta_{\mathrm{defl,min}}=45$ for $e_0=0.90$ \drafting{need to insert degrees sign, but command not working?}.
To quantify the eccentricity of the resulting SMBH binary, we determine when the binary becomes hard (\autoref{eq:ahard}), and calculate the median and interquartile range (IQR) over a maximum of ten orbits about the time when $t=t(a=a_\mathrm{h})$.
These points are shown in \drafting{FIGURE}, where the IQR is generally smaller than the symbol marker used in the plot.

\begin{figure}
    \includegraphics{orbit}
    \caption{\drafting{Orbit of a SMBH in one of the $e_0=0.99$ mergers, which by symmetry of the equal mass system is the same as the second BH orbit reflected. Line coloured according to time. Point A shows a time just before the first hard scattering, and point B a time just after. The circle marker indicates the final position of the BH before the BHs become bound.}}
    \label{fig:orbit}
\end{figure}

>>>>

\section{Results}\label{sec:results}

We present our findings first for the SMBH binary eccentricity scatter for initially low-eccentricity orbits (\autoref{ssec:lowe}), and then present the findings for initially high-eccentricity orbits (\autoref{ssec:highe}).
We explain the results using an analytical representation in \autoref{ssec:toy}.

@@

\subsection{Eccentricity scatter in low-e orbits}\label{ssec:lowe}

We show the dependence of eccentricity standard deviation on mass resolution for the $e_0=0.90$ orbit in \autoref{fig:convergence} with purple markers. 
Taking the full set of ten simulations for this set (circle markers), immediately apparent is the lack of convergence of $\sigma_e$, with $5\times10^3$ and $1\times10^4$ mass resolutions demonstrating an increased scatter compared to the lowest mass resolution of $2.5\times10^3$.
The peculiar values of $\sigma_e$ arises due to the assumed Gaussian distribution of the eccentricity between simulations.
Inspecting the orbital parameters for all simulations that form the $e_0=0.90$ suite, it is found that for resolutions of $5\times10^3$ and $1\times10^4$ the eccentricity clusters around a value of $e_\mathrm{h}\simeq0.90$, except for one simulation at each resolution which has a value of $e_\mathrm{h}<0.4$.
This one outlier for each of the two resolutions artificially inflates the measured eccentricity variance, and by applying a $2.5\sigma$-cut in eccentricity we recover the $1/\sqrt{M_\bullet/m_\star}\propto 1/\sqrt{N_\star}$ scaling of $\sigma_e$ as in \citet{nasim2020} (square markers), apart from the highest mass resolution set of $M_\bullet/m_\star=4\times10^4$.
However, the eccentricity values beyond $2.5\sigma$ from the mean are each a valid measurement from the simulation, and highlight the presence of stochasticity in eccentricity for these idealised simulation set-ups. 
The cause of these low eccentricity values is discussed in \autoref{ssec:toy}.

\subsection{Eccentricity scatter in high-e orbits}\label{ssec:highe}

We show the dependence of eccentricity standard deviation on mass resolution for the $e_0=0.99$ orbit in \autoref{fig:convergence} with blue markers, where the marker symbols are consistent with those for the $e_0=0.90$ suite. 
We find that the scaling of $\sigma_e$ is steeper than the $1/\sqrt{M_\bullet/m_\star}\propto1/\sqrt{N_\star}$ relation described in \citet{nasim2020}.
We attribute this dramatic decrease in $\sigma_e$ to an inflated value of $\sigma_e$ at low mass resolution: there is increased scatter in eccentricity at low mass resolutions for very radial initial orbits of the SMBH binary. 
This observation is highlighted in \autoref{fig:nasim_e_comp}.
As the majority of major galaxy mergers in the real Universe are expected to be radial \citep[e.g.][]{khochfar2006}, understanding if the origin of the scatter in eccentricity is a numerical or a physical effect is of paramount importance, and will have direct implications for the predictions of SMBH merger rates by ongoing and future GW detection missions.

@@

After specifying the initial position and velocity, as well as the other parameters of the model,
the equation of motion can be solved using standard numerical integration techniques.
We use the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{virtanen2020}
for this purpose.
To mimic the nearly linearly plunging orbit in the high-eccentricity merger orbit simulations,
we specify the initial conditions confined to the $xz$-plane as 
\begin{gather}
\vb{x}_0 = (r_0, b)\\
\dot{\vb{x}}_0 = (-v_0, 0).
\end{gather}
We fix $r_0 = 100\,\mathrm{pc}$,
and use model parameters
$\rho = 400 \,\Msun\,\mathrm{pc}^{-3}$,
$\sigma=200\,\mathrm{km\,s^{-1}}$,
$M_\bullet=10^8\,\Msun$,
and
$e_\mathrm{s}=0.85$,
which match fairly well with the central properties of the merging galaxy when
the SMBHs are becoming bound in the $e_0=0.99$ merger orbit simulations.
\drafting{(Check the params)}

\drafting{This right place for this section?}
To investigate the effect of the deflection angle, and by extension the impact parameter, on the final SMBH binary eccentricity, we determine the deflection angle at each pericentre passage before the SMBHs become bound \citep{binney2008}:
\begin{equation}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, and $L$ is the magnitude of the SMBH binary angular momentum vector and $E$ the binary orbital energy at the time of the pericentre passage.
We set as the first `hard-scattering' angle $\theta_\mathrm{defl}$ the first angle that exceeds an orbit-dependent threshold: $\theta_{\mathrm{defl,min}}=90$ for $e_0=0.99$ and $\theta_{\mathrm{defl,min}}=45$ for $e_0=0.90$ \drafting{need to insert degrees sign, but command not working?}.
An example of $\theta_\mathrm{defl}$ is shown for two realisations of the $e_0=0.99$ orbit in \autoref{fig:orbit}, which occurs between the points A and B in the SMBH binary orbit in each inset panel. 
To quantify the eccentricity of the resulting SMBH binary, we determine when the binary becomes hard (\autoref{eq:ahard}), and calculate the median and interquartile range (IQR) over a maximum of ten orbits about the time when $t=t(a=a_\mathrm{h})$.
These points are shown in \drafting{FIGURE}, where the IQR is generally smaller than the symbol marker used in the plot.

\begin{figure}
    \includegraphics{orbit}
    \caption{Orbits from $t=0\,\mathrm{Myr}$ to $t=t_\mathrm{bound}$ of a single SMBH in two realisations of the $e_0=0.99$ mergers, which by symmetry of the equal mass system is the same as the second BH orbit reflected. Each line is coloured according to time of the simulation, however the colour for $t<22\,\mathrm{Myr}$ is constant for visual clarity of the later evolution. In each inset panel, point A shows a time just before the first hard scattering $\theta_\mathrm{defl}$, and point B a time just after. The circle marker indicates the final position of the SMBH before the SMBHs form a bound binary.}
    \label{fig:orbit}
\end{figure}
